Two applications of logic programming to Coq

### 2020-11-05 papers/types2020/cultures.tex

Given that \lP and Coq both result from combining the
$\lambda$-calculus with logic, it is important to understand some of
their their differences.  The confusion around the term
\emph{higher-order abstract syntax} (HOAS), is a case in point.  In
the functional programming setting, including the Coq system, the HOAS
approach leads to encoding binding structures within terms using
functions.  The earliest such encodings were unsatisfactory since they
would allow for \emph{exotic terms} \cite{despeyroux95tlca} and for
structures on which induction was not possible \cite{roeckl01fossacs}.
Later approaches yielded non-canonical and complex encodings, although
these could support inductive and coinductive reasoning
\cite{chlipala08icfp,honsell01tcs,pientka10ijcar}. 
% DM I'm hoping that the additional sentence and references are enough
% here. 
%% \begin{metanote}
%%   AM: I find this a bit unbalanced. Exotic terms can be avoided with
%%   Honsell's theory of context and/or Adam's PHOAS -- not that I like
%%   it.. And of course there is Beluga and friends.  On the other hand
%%   Coq users are happy with library-based approaches such as Autosubt
%%   or well-typed DB that allows for sphisticated treatments of
%%   binders. I would say something more generic such as that mixing
%%   functions and HOAS is not immediate, may require additional
%%   machinery both in terms of type theories than of code generation,
%%   while in lp\dots
%% \end{metanote}
In the logic programming setting, particularly \lP, HOAS is well
supported since bindings are allowed with terms ($\lambda$-bindings),
in formulas (quantifiers), and in proofs (eigenvariables).  (In fact,
the original paper on HOAS \cite{pfenning88pldi} was inspired by \lP.)
For this reason, the term \emph{$\lambda$-tree syntax} was introduced
to name this particular take on HOAS \cite{miller19jar}.  The Abella
proof assistant \cite{baelde14jfr} was designed, in part, to provide
inductive and co-inductive inference involving specifications using
the $\lambda$-tree syntax approach.

>>>>

Given that \lP and Coq both result from combining the
$\lambda$-calculus with logic, it is important to understand some of
their their differences.  The confusion around the term
\emph{higher-order abstract syntax} (HOAS), is a case in point.  In
the functional programming setting, including the Coq system, the HOAS
approach leads to encoding binding structures within terms using
functions.  The earliest such encodings were unsatisfactory since they
would allow for \emph{exotic terms} \cite{despeyroux95tlca} and for
structures on which induction was not immediately possible \cite{roeckl01fossacs}.
Later approaches yielded non-canonical and complex encodings~\cite{chlipala08icfp,honsell01tcs}, as well as sophisticated type theories~\cite{pientka10ijcar}. All of
these could support inductive and coinductive reasoning. 
% DM I'm hoping that the additional sentence and references are enough
% here. 
%% \begin{metanote}
%%   AM: I find this a bit unbalanced. Exotic terms can be avoided with
%%   Honsell's theory of context and/or Adam's PHOAS -- not that I like
%%   it.. And of course there is Beluga and friends.  On the other hand
%%   Coq users are happy with library-based approaches such as Autosubt
%%   or well-typed DB that allows for sphisticated treatments of
%%   binders. I would say something more generic such as that mixing
%%   functions and HOAS is not immediate, may require additional
%%   machinery both in terms of type theories than of code generation,
%%   while in lp\dots
%% \end{metanote}
In the logic programming setting, particularly \lP, HOAS is well
supported since bindings are allowed with terms ($\lambda$-bindings),
in formulas (quantifiers), and in proofs (eigenvariables).  (In fact,
the original paper on HOAS \cite{pfenning88pldi} was inspired by \lP.)
For this reason, the term \emph{$\lambda$-tree syntax} was introduced
to name this particular take on HOAS \cite{miller19jar}.  The Abella
proof assistant \cite{baelde14jfr} was designed, in part, to provide
inductive and co-inductive inference involving specifications using
the $\lambda$-tree syntax approach.

### 2020-11-05 papers/types2020/intro.tex

Recently, C. Sacerdoti Coen and E. Tassi developed the Elpi
implementation \cite{dunchev15lpar} of \lP \cite{miller12proghol} and
more recently, Tassi has made Elpi available as the \texttt{coq-elpi}
plugin to the Coq proof assistant \cite{tassi18coqpl}
(\url{https://github.com/LPCIC/coq-elpi}).  This implementation of \lP
extends earlier ones in primarily two directions: First, Elpi adds a
notion of constraints and constraints handling rules, which makes it
more expressive than the Teyjus implementation \cite{nadathur99cade}
of \lP. Second, the plugin version of Elpi comes equipped with a
quotation and anti-quotation syntax for mixing Coq expressions with
\lP program elements and an API for accessing the Coq environment,
type checking included.

@@

Felty has also made the point that LCF-style tactics and tacticals can
be given an elegant and natural specification using the higher-order
relational specifications provided by \lP \cite{felty93jar}.  Some
recent implementations built using Elpi support the usefulness of
higher-order logic programming as a meta-programming language for
proof assistants in general~\cite{DunchevCT16,coen19mscs} and, in
particular, for Coq via the \texttt{coq-elpi} plug
in~\cite{CohenST20,tassi19itp} .

>>>>

Recently, C. Sacerdoti Coen and E. Tassi developed the Elpi
implementation \cite{dunchev15lpar} of \lP \cite{miller12proghol} and
more recently, Tassi has made Elpi available as the Coq-Elpi
plugin~\cite{tassi18coqpl} (\url{https://github.com/LPCIC/coq-elpi})
to the Coq proof assistant.  This implementation of \lP extends
earlier ones in primarily two directions: First, Elpi adds a notion of
constraints and constraints handling rules, which makes it more
expressive than the Teyjus implementation \cite{nadathur99cade} of
\lP. Second, the plugin version of Elpi comes equipped with a
quotation and anti-quotation syntax for mixing Coq expressions with
\lP program elements and an API for accessing the Coq environment,
type checking included.

@@

Felty has also made the point that LCF-style tactics and tacticals can
be given an elegant and natural specification using the higher-order
relational specifications provided by \lP \cite{felty93jar}.  Some
recent implementations built using Elpi support the usefulness of
higher-order logic programming as a meta-programming language for
proof assistants in general~\cite{DunchevCT16,coen19mscs} and, in
particular, for Coq via the Coq-Elpi plug~\cite{CohenST20,tassi19itp} .

### 2020-11-05 papers/types2020/sec3.tex

\newpage
\section{Proof theory and proof certificates}
\label{sec:three}
  \newcommand{\XXi}{{\color{blue}{\Xi}}}

@@

%\begin{footnotesize} % No effect on typesetting
\begin{figure}[t]
  %
%  \renewcommand{\Gamma}{}
    \newcommand{\bc}[5]{#1#2\Downarrow #3 : #4 \vdash #5}
\[
% \infer{\XXi\vdash t = t}
% {%\eqExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash \true}
% {%\trueExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash G_1\vee G_2}
% {\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
% }
% \qquad
% \infer{\XXi\vdash \exists x. G}
% {\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
% }
% \qquad 
% \qquad
% \infer{\XXi;\Gamma\vdash \lambda x:A.\,t : \forall x : A.\,  G}
% {\XXi;\Gamma, x : A\vdash t : G%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
% }
%\qquad
\infer{\XXi_1\vdash k\ l : A}
      {\mathrm{Ind}[p] \ (\Gamma_I := \Gamma_C) \in E\quad (\mathit{head} \ A : T)\in \Gamma_I\quad  k : D \in \Gamma_C\quad \bc{\XXi_2}{} l D A \quad\unfoldExpert{\XXi_1}{\XXi_2}} 
\]
\vskip -18pt
\[
\infer{\XXi\vdash  s : A}
{%(A,s) \in \mathcal{A}
  {E[] \vdash_{CIC} s : A}% The order should be s : A, not A : s ?
}
\qquad
  \infer{\bc {\XXi}{} {[]} A A}{\trueExpert{\XXi}}{}\qquad
%  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
  \infer{\bc{\XXi}{}{(t\ ::\ l)} {\forall x:B.\, D} A}
  {\XXi_1\vdash t : B\qquad \bc{\XXi_2}{}l {D[t/x]} A\qquad\prodExpert{\XXi}{\XXi_1}{\XXi_2}}{}
\]
%\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
\caption{Specification of the checking of a proof certificate and the
  synthesis of a dependently typed $\lambda$-term.}
\label{fig:augmented}
\end{figure}
\begin{figure}
\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={sigs-end}]{code/dep-kernel.sig}
\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={check-end}]{code/dep-kernel-v2.mod}
\caption{Implementation of a  dependent kernel for Coq terms.}
\label{fig:kernel}
\end{figure}
%\end{footnotesize}

@@

Coq terms are accessed through the Coq-Elpi API, and their representation
in \lP takes advantage of the native \lP constructs such as lists and binders.
The following is a list of some of signatures of the notions that we use:
\begin{lstlisting}
  kind term   type.                                   % reification of Coq terms
  kind gref    type                                 % reification of refernces to global terms
  type global gref -> term.                      % coercion to term
  type indt   inductive -> gref.                 % reification of inductive types
  type indc   constructor -> gref.             % reification of their constructors 
  type app    list term -> term.               % reification of nary application
  type prod   name -> term -> (term -> term) -> term. % reification of dependent product
\end{lstlisting}

Note that \lsti{prod} encodes dependent products taking a name for pretty printing,  a term and a \lP function from
terms to terms: namely $\forall x : A, B$ is encoded by \texttt{prod ``x'' A (x\ B x)};
when, in the implementation of the product-left rule, we
apply $B$ to the variable \lsti{Tm},
we get a term that can be used to continue backchaining. This
application is obtained via meta-level substitution, in the style of HOAS. In this sense, our calculus uses \emph{implicit} substitutions, rather than explicit ones as in LJT and PTSC's tradition, the more since proof search in our application is cut-free. 
% Syntax: one can think as if something of type \lsti{term -> term} were a
% \lsti{term} with a hole in it for another \lsti{term}.
\todo{Reput safe-dest-app in code, a bit more details about the unfold case}
The unfolding rule makes use of the Coq-Elpi primitives
\lsti{coq.safe-dest-app} in order to obtain the head term of a (possibly
nested) application, and \lsti{coq.env.indt} in order to access the global
environment of inductive definitions and query for information about them.

In Figure~\ref{fig:augmented}  each
inference rule is augmented with an additional premise involving an
\emph{expert predicate}, a certificate $\Xi$, and possibly continuations of
certificates ($\Xi'$, $\Xi_1$, $\Xi_2$) with extracted information from
These notions are taken from the general setting of \emph{foundational
  proof certificates}~\cite{chihani17jar}.
%
In our case here, an FPC is a collection of \lP clauses that
provide the remaining details not supplied in Figure~\ref{fig:augmented} and~\ref{fig:kernel}:
that is, the exact set of constructors for the  type of certificates as
well as the specification of the expert predicates listed \emph{ibidem}.
%
The top of Figure~\ref{fig:resources} displays two FPCs,
both of which can be used to describe proofs where we bound
the dimension of  a proof.
%
For example, the first FPC dictates that the query \mbox{\lsti{(check (qheight 5) B)}} is
provable in  the kernel using the clauses in Figures~\ref{fig:kernel}
and~\ref{fig:resources} if and only if the height of that proof is 5
or less.
%
Similarly, the second FPC can be used to bound the total number of instances of
unfoldings in a proof.
%
%In particular, the query \mbox{\lsti{(check (qsize 5 H) B)}}
% \begin{lstlisting}
% sigma H\ (check (qsize 5 H) B)
% \end{lstlisting}
%is provable if and only if the total number of unfoldings of that
%proof is 5 or less.

>>>>

\newpage
\section{Proof theory and proof certificates}
\label{sec:three}

\newcommand{\XXi}{{\color{blue}{\Xi}}}
\newcommand{\bc}[5]{#1#2\Downarrow #3 : #4 \vdash #5}

@@

\begin{figure}[t]
\[
\infer{\XXi_1\vdash k\ l : A}
      { \mathrm{Ind}[p] \ (\Gamma_I := \Gamma_C) \in E\quad 
       (\mathit{head} \ A : T)\in \Gamma_I\quad  
       k : D \in \Gamma_C\quad 
       \bc{\XXi_2}{} l D A \quad\unfoldExpert{\XXi_1}{\XXi_2}} 
\]
\vskip -18pt
\[
  \infer{\XXi\vdash  s : A}{{E[] \vdash_{CIC} s : A}}% The order should be s : A, not A : s ?
  \qquad
  \infer{\bc{\XXi}{}{[]} A A}{\trueExpert{\XXi}}
  \qquad
  \infer{\bc{\XXi}{}{(t\ ::\ l)}{\forall x:B.\, D} A}
        {\XXi_1\vdash t : B\qquad
         \bc{\XXi_2}{}l {D[t/x]} A\qquad
         \prodExpert{\XXi}{\XXi_1}{\XXi_2}}
\]
\caption{Specification of the checking of a proof certificate and the
  synthesis of a dependently typed $\lambda$-term.}
\label{fig:augmented}
\end{figure}

\[
  \infer{\Pscr\vdash A}{D \in \Pscr\quad \Pscr\Downarrow D\vdash A}
  \qquad
  \infer{\Pscr\Downarrow A\vdash A}{}
  \qquad
  \infer{\Pscr\Downarrow\forall x. B(x)\supset D(x)\vdash A}
        {\vdash B(t)\qquad \Pscr\Downarrow D(t)\vdash A}
\]
Here, $\Pscr$ is a collection of Horn clauses, i.e., formulas of the
form $\forall \bar x_1. A_1\supset \forall \bar x_2. A_2\supset
\forall \bar x_n. A_n\supset A_0$.

\begin{figure}
\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={sigs-end}]{code/dep-kernel.sig}
\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={check-end}]{code/dep-kernel-v2.mod}
\caption{Implementation of a  dependent kernel for Coq terms.}
\label{fig:kernel}
\end{figure}

@@

Coq terms are accessed through the Coq-Elpi API, and their representation
in \lP takes advantage of the native \lP constructs such as lists and binders.
The following is a list of some of the signatures of the notions that we use:
\begin{lstlisting}
  kind term   type.                                   % reification of Coq terms
  kind gref    type                                 % reification of refernces to global terms
  type global gref -> term.                      % coercion to term
  type indt   inductive -> gref.                 % reification of inductive types
  type indc   constructor -> gref.             % reification of their constructors 
  type app    list term -> term.               % reification of nary application
  type prod   name -> term -> (term -> term) -> term. % reification of dependent product
\end{lstlisting}

Note that \lsti{prod} encodes dependent products taking a name for pretty
printing, a term and a \lP function from terms to terms: namely $\forall x :
A, B$ is encoded by \texttt{prod ``x'' A (x\textbackslash{} B x)}; when, in the
implementation of the product-left rule, we apply $B$ to the variable
\lsti{Tm}, we get a term that can be used to continue backchaining. This
application is obtained via meta-level substitution, in the style of HOAS. In
this sense, our calculus uses \emph{implicit} substitutions, rather than
explicit ones as in LJT and PTSC's tradition, the more since proof search in
our application is cut-free.
% Syntax: one can think as if something of type \lsti{term -> term} were a
% \lsti{term} with a hole in it for another \lsti{term}.
The unfolding rule makes use of the Coq-Elpi primitives
\lsti{coq.safe-dest-app} in order to obtain the head term of a (possibly
nested) application, and \lsti{coq.env.indt} in order to access the global
environment of inductive definitions and query for information about them.
The \lsti{unfold_expert} predicate, then, has the role of selecting which
constructor to introduce for the inductive type. The kernel will successively
obtain the type of the selected constructor, and initiate the backchaining
phase on it.

In figure~\ref{fig:augmented}  each
inference rule is augmented with an additional premise involving an
\emph{expert predicate}, a certificate $\Xi$, and possibly continuations of
certificates ($\Xi'$, $\Xi_1$, $\Xi_2$) with extracted information from
These notions are taken from the general setting of \emph{foundational
  proof certificates}~\cite{chihani17jar}.
%
\todo{MM: Talk about randomized FPC?}
In our case here, an FPC is a collection of \lP clauses that
provide the remaining details not supplied in Figure~\ref{fig:augmented} and~\ref{fig:kernel}:
that is, the exact set of constructors for the  type of certificates as
well as the specification of the expert predicates listed \emph{ibidem}.
%
The top of Figure~\ref{fig:resources} displays two FPCs,
both of which can be used to describe proofs where we bound
the dimension of  a proof.
%
For example, the first FPC dictates that the query \mbox{\lsti{(check (qheight 5) B)}} is
provable in  the kernel using the clauses in Figures~\ref{fig:kernel}
and~\ref{fig:resources} if and only if the height of that proof is 5
or less.
%
Similarly, the second FPC can be used to bound the total number of instances of
unfoldings in a proof.
%
%In particular, the query \mbox{\lsti{(check (qsize 5 H) B)}}
% \begin{lstlisting}
% sigma H\ (check (qsize 5 H) B)
% \end{lstlisting}
%is provable if and only if the total number of unfoldings of that
%proof is 5 or less.
